---
title: "Google Gauge Monitoring - Review"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
    embed-resoures: true
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
  eval: true
  results: "asis"
  out.width: "100%"
editor: visual
project:
     execute-dir: project
---

## Intro

This document reviews the gauge discharge (Q) value obtained from Gooles flood forecasting googlesheet. It also looks at the values with respect to the return period values calculated (per gauge) by google and our basin level threshold.

For reference the currently alert level is set at : ≥ 80 % of gauges in a basing breaching the 2 year RP (in the next 7 days)


```{r}
GAUGES_TO_REMOVE <- c(
  "hybas_1120794570",
  "hybas_1120741070",
  "hybas_1120946640",
  "hybas_1120974450",
  "hybas_1120981190"
)
# install_github("r-tmap/tmap@v4")
# libs --------------------------------------------------------------------
# in dedicated monitoring repo should consider setting up w/ {renv}
# have to load tidyverse packages separate for GHA for some reason
library(dplyr)
library(purrr)
library(readr)
library(here)
library(ggplot2)
library(tidyr) # pivot wider
library(stringr)
library(lubridate)
library(sf)
library(googlesheets4)
library(janitor)
library(tmap)
library(here)
library(blastula)
library(googledrive)
library(ggtext) # colored title
library(glue)
library(gghdx)
library(rhdx)
gghdx()

source(here(file.path("R","email_funcs.R")))
source(here(file.path("src","email","email_utils.R")))
```


```{r gaugeSnapshotCreation, eval = F}
# Get Data ----------------------------------------------------------------

## Authenticate google APIs ####
walk(list(drive_auth, gs4_auth), \(f){
  f(
    path = Sys.getenv("GFF_JSON")
  )
})

drive_dribble <- drive_ls(
  corpus = "user"
)


## Static Layers ####
# load layers for mapping from hdx
L <- hdx_map_viz_layers()

# Basins
drive_download(
  as_id("1AsB_Cf9QQb3vkp9ebIFvTPZ5gBYqYRJR"),
  path = basin_fp <- tempfile(fileext = ".rds")
)

gdf_basins_poly <- read_rds(basin_fp) %>%
  filter_basins() %>%
  # clip to country
  st_intersection(
    L$west_central_africa %>%
      filter(admin0Pcod == "NG")
  )

## Live Data ####
### Google forecast Workbook ####
gauge_df_list <- read_gauge_googlesheets(
  url = Sys.getenv("GFF_GAUGE_URL")
)


dff <- gauge_df_list %>%
  keep_at(at = ~ str_detect(.x, "hybas_")) %>%
  bind_rows() %>%
  # remove potential duplication issue where two discharge values may get appended to googlesheed w/ same date
  mutate(
    date = dmy(date),
    update_time_utc = ymd_hms(update_time_utc)
  ) %>%
  group_by(gauge_id,date) %>%
  filter(
    update_time_utc == max(update_time_utc) # get max update time (in case duplicates)
  ) %>%
  ungroup() %>%
  filter(
    !gauge_id %in% GAUGES_TO_REMOVE
  ) %>%
  # add basin informatino to gauge/forecast data
  left_join(
    gdf_gauge %>%
      st_drop_geometry(),
    by = "gauge_id"
  ) %>%
  pivot_longer(
    cols = matches("discharge"),
    names_to = "forecast_cat",
    values_to = "Q"
  ) %>%
  mutate(
    leadtime = replace_na(parse_number(forecast_cat), 0) %>%
      suppressWarnings(), # don't need the parse_number warning
    date_predict = date + leadtime,
    Q_pct_rp2 = Q / x2_years_return_period,
    gte_2_rp = Q >= x2_years_return_period,
    gte_5_rp = Q >= x5_years_return_period,
    gte_20_rp = Q >= x20_years_return_period
  ) 


write_csv(
  dff,
  file.path(Sys.getenv("GFH_DATA_DIR"),"inputs","20231113_google_gauge_snapshot.csv") 
)

```



```{r}
dff <- read_csv(file.path(Sys.getenv("GFH_DATA_DIR"),
                          "inputs","20231113_google_gauge_snapshot.csv") )

dff2 <- dff %>% 
  group_by(
    basin_name, gauge_id, date
  ) %>% 
  summarise(
    Q= max(Q),
    across(
      .cols = starts_with("x"),
      .fns = ~unique(.x)
    ),.groups="drop"
  ) %>% 
  mutate(
    gte_rp2 = Q>x2_years_return_period,
    gte_rp5 = Q>x5_years_return_period,
    gte_rp20 = Q>x20_years_return_period
  ) 
```


## Individual Gauge Discharge

```{r}
dff2$gauge_id %>% unique() %>% length()
dff2 %>% 
  group_by(gauge_id) %>%
  mutate(
    flag2 = if_else(any(gte_rp2),"breached 2 year RP","No Breach")
  ) %>%
  ungroup() %>%
  ggplot(
    aes(x= date, y= Q, group=gauge_id,color=flag2)
  )+
  geom_line()+
  facet_wrap(
    ~basin_name
  )
dff2 %>% 
  group_by(gauge_id) %>%
  mutate(
    flag5 = if_else(any(gte_rp5),"breached 5 year RP","No Breach")
  ) %>%
  ungroup() %>%
  ggplot(
    aes(x= date, y= Q, group=gauge_id,color=flag5)
  )+
  geom_line()+
  facet_wrap(
    ~basin_name
  )
dff2 %>% 
  group_by(gauge_id) %>%
  mutate(
    flags = case_when(
      any(gte_rp20)~"breached 20 year RP",
      any(gte_rp5)~"breached 5 year RP",
      any(gte_rp2)~"breached 2 year RP",
      TRUE~"No Breach"
    )
  ) %>%
  ungroup() %>%
  ggplot(
    aes(x= date, y= Q, group=gauge_id,color=flags)
  )+
  geom_line()+
  scale_color_manual(values = c(
    # "breached 20 year RP" = hdx_hex("tomato-hdx"),
    "breached 5 year RP" = hdx_hex("tomato-hdx"),
    "breached 2 year RP" = hdx_hex("sapphire-hdx"),
    "No Breach" = hdx_hex("mint-hdx")
  ))+
  facet_wrap(
    ~basin_name
  )
```

## % Gauges Breaching per Basin
```{r}

dff2_summarised <- dff2 %>% 

  group_by(basin_name,date) %>% 
  summarise(
    across(.cols =starts_with("gte"),
           .fns= list(sum = sum, pct = mean),
           .names = "rp_{parse_number(col)}_{fn}"),.groups="drop"
    ) 
  
dff2_summarised %>% 
  ggplot(
    aes(x= date, y=  rp_2_pct, group=basin_name, color=basin_name)
  )+
  geom_line()+
  scale_y_continuous(labels=scales::label_percent(), limits = c(0,1))+
  geom_hline(
    yintercept = 0.8, linetype="dashed", color=hdx_hex("tomato-hdx")
  )+
  labs(
    y= "% gauges breaching 2 year RP",
    title ="% Gauges breaching 2 year RP per basin over time",
    caption= "Lower Niger Basin sustained high a high % of gauges breaching 2 year RP (10 out of 15), but never breached 80 % threshold"
  )+
  theme(
    legend.title = element_blank(),
    axis.title.x=element_blank()
  )


dff2_summarised %>% 
  ggplot(
    aes(x= date, y=  rp_5_pct, group=basin_name, color=basin_name)
  )+
  geom_line()+
  # scale_y_continuous(labels=scales::label_percent(), limits = c(0,1))+
  scale_y_continuous(labels=scales::label_percent())+
  # geom_hline(
  #   yintercept = 0.8, linetype="dashed", color=hdx_hex("tomato-hdx")
  # )+
  labs(
    y= "% gauges breaching 2 year RP",
    title ="% Gauges breaching 2 year RP per basin over time",
    caption= "Lower Niger Basin sustained high a high % of gauges breaching 5 year RP (10 out of 15), but never breached 80 % threshold"
  )+
  theme(
    legend.title = element_blank(),
    axis.title.x=element_blank()
  )

```

## Number Gauges Breaching per Basin
```{r}
  dff2_summarised %>% 
  ggplot(
    aes(x= date, y=  rp_2_sum, group=basin_name, color=basin_name)
  )+
  geom_line()+
  scale_y_continuous()+
  labs(
    y= "% gauges breaching 2 year RP",
    title ="% Gauges breaching 2 year RP per basin over time",
    caption= "Lower Niger Basin sustained 10/15 gauges breaching threshold for most of September"
  )+
  theme(
    legend.title = element_blank(),
    axis.title.x=element_blank()
  )


```
